---
editor_options: 
  markdown: 
    wrap: 128
---

# Memory and Register Declaration

A 32 bit integer is used as the number of memory locations for the VM. Given that the max value of a signed 16 bit integer is
*65536* or (2\^16), that is how many data storage locations are present in the VM. There are additionally 19 registers of 32 bit
integers, most of which (16) are general purpose and can be used to send data to the emulated CPU. The memory locations and
registers were defined and stored as an array.

"The general purpose registers can be used to perform any program calculations. The program counter is an unsigned integer which
is the address of the next instruction in memory to execute. The condition flags tell us information about the previous
calculation." - an excerpt from (<https://www.jmeiners.com/lc3-vm/#what-is-a-virtual-machine->).

```{r}
MEMORY_MAX = 2^16 # 128KB allocated as memory locations (32 bits each)
MemoryList <- integer(MEMORY_MAX)

Registers <- list (  
    R0      = 1, 
    R1      = 2, 
    R2      = 3, 
    R3      = 4, 
    R4      = 5,
    R5      = 6, 
    R6      = 7, 
    R7      = 8,
    PC      = 9, 
    COND    = 10, 
    COUNT   = 11 
)

MMR <- c(
    MR_KBSR = 0xFE00, # keyboard status
    MR_KBDR = 0xFE02  # keyboard data
)

mem_write <- function(address, val) {
    MemoryList[address] = val
}

mem_read <- function(address) {
    if (address == MMR$MR_KBSR) {
        if (check_key()) {
            MemoryList[MMR$MR_KBSR] <- bitwShiftL(1, 15)  
             character
            MemoryList[MMR$MR_KBSR] <- as.integer(charToRaw(readline("Enter a character: "))[1])
        } else {
            MemoryList[MMR$MR_KBSR] <- 0
        }
    }
    
    return(MemoryList[address])

}

RegisterValList <- integer(length = 19) # 19 = number of enumerated values
```

# Opcodes and Flags

An opcode is a command that describes which action the CPU should carryout. CPU instructions have both an opcode and parameters,
the opcode states which command should be carried out and the parameters act as inputs to the task being performed.

The COND register stores condition flags, in this instance, that condition exclusively refers to whether a given value is
positive or negative. By bitshifting each value by a given amount, we can implement this condition flag feature.

```{r}
Opcodes <- list (
    BR    = 1,    # branch
    ADD   = 2,    # add
    LD    = 3,    # load 
    ST    = 4,    # store
    JSR   = 5,    # jump register 
    AND   = 6,    # bitwise and 
    LDR   = 7,    # load register
    STR   = 8,    # store register
    RTI   = 9,    # unused 
    NOT   = 10,    # bitwise not
    LDI   = 11,   # load indirect
    STI   = 12,   # store indirect
    JMP   = 13,   # jump
    RES   = 14,   # reserved (unused)
    LEA   = 15,   # load effective address
    TRAP  = 16    # execute trap
)

Flags <- list (
    FLAG_ZERO       = bitwShiftL(1, 0),   # Zero
    FLAG_POSITIVE   = bitwShiftL(1, 1),   # Postive
    FLAG_NEGATIVE  = bitwShiftL(1, 2)    # Negative
)
FlagList <- integer(length(Flags))

```

# Helper Functions for Instructions

```{r}

To16Bit <- function(x) {
  # Check if x is NULL or empty
  if (is.null(x) || length(x) == 0) {
    cat("To16Bit received NULL or empty value\n")
    return(0)  # Return 0 as a default fallback
  }
  
  x <- bitwAnd(x, 0xFFFF)
  
  # Debug print to check x after masking
  cat("Value after masking to 16-bit: ", x, "\n")
  
  if (x > 0x7FFF) {
    x <- x - 0x10000  # Two's complement
  }
  
  return(x)
}

GetBitCount <- function(x) {
  # If x is 0, return 1 since we need at least 1 bit to represent zero
  if (x == 0) {
    return(1)
  }
  
  # Calculate the number of bits required to represent x
  return(floor(log2(abs(x))) + 1)
}

SignExtend <- function(x, BitCount) {
  # Debugging check
  cat("Original value before SignExtend: ", x, "\n")
  
  x <- To16Bit(x)  # Ensure it's a 16-bit value
  # Check if the sign bit (BitCount - 1) is set
  if (bitwShiftR(x, BitCount - 1) & 1) {
    # If set, extend the sign by setting the upper bits
    x <- bitwOr(x, bitwShiftL(0xFFFF, BitCount))
  }
  cat("Value after SignExtend: ", x, "\n")
  
  return(x)
}

# UpdateFlags function to check and set flags based on register value
UpdateFlags <- function(R) {
  # Check if the input register value is NULL or empty
  RBits <- GetBitCount(R)
  RExtended <- SignExtend(R, RBits)  # Call SignExtend with R and RBits
  
  if (is.null(RExtended) || length(RExtended) == 0) {
    return(NULL)
  }
  
  # Check the value and set the condition flag directly
  if (RExtended == 0) {
    RegisterValList[Registers$COND] <- Flags$FLAG_ZERO  # Set the zero flag (1)
  } else if ((bitwAnd(RExtended, 0x8000)) != 0) {  # Check if the sign bit (MSB) is 1 (negative)
    RegisterValList[Registers$COND] <- Flags$FLAG_NEGATIVE  # Set the negative flag (4)
    
  } else {
    RegisterValList[Registers$COND] <- Flags$FLAG_POSITIVE  # Set the positive flag (2)
    
  }
}

```

# Main Loop

```{r}

RegisterValList[Registers$COND] <- Flags$FLAG_ZERO #apply FLAG_ZERO to the COND register

PC_START <- 0x3000 #set computer to default value at start
RegisterValList[Registers$PC] <- PC_START

mem_read_value <- function(address) {
  return(MemoryList[address]) #add 1 if not working as intended
}

handle_interupt <- function() {
    cat("\nInterrupt signal (SIGINT) received. Exiting gracefully.\n")
    quit(save = "no", status = 1)
}
setInterruptHandler(handle_interrupt)

disable_input_buffering();

signal(SIGINT, handle_interrupt);
disable_input_buffering();

running <- TRUE;
while (running) {
    Instruction <- mem_read_value(RegisterValList[Registers$PC])
    RegisterValList[Registers$PC] <- (RegisterValList[Registers$PC])

    OperationCode <- bitwShiftR(Instruction, 12)

    switch (OperationCode,
    ADD = ADD(),
    LDI = LDI(),
    RTI = RTI(),
    AND = AND(),
    NOT = NOT(),
    BR = BR(),
    JMP = JMP(),
    JSR = JSR(),
    LD = LD(),
    LDR = LDR(),
    LEA = LEA(),
    ST = ST(),
    STI = STI(),
    STR = STR(),
    RES()        #default for bad code
)
    restore_input_buffering();
}

```

# Instruction Functions

So cool it rhymes!

```{r}

ADD <- function() {
  RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)   # destination register
  RegisterValList[Registers$R1] <- bitwAnd(bitwShiftR(Instruction, 6), 0x7) # first operand
  ImmFlag <- bitwAnd(bitwShiftR(Instruction, 5), 0x1) #Immediate Mode Flag
  
  if (ImmFlag) {
    imm5 <- SignExtend(bitwAnd(Instruction, 0x1F), 5)
    RegisterValList[Registers$R0] = RegisterValList[Registers$R1] + imm5
    
  } else {
    r2 <- bitwAnd(Instruction, 0x7)
    RegisterValList[Registers$R0] = RegisterValList[Registers$R1] +     RegisterValList[Registers$R2]
  }
  UpdateFlags(RegisterValList[Registers$R0])
}

LDI <- function() {
  RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
  PCOffset <- SignExtend(bitwAnd(Instruction, 0x1FF), 9)
  
  RegisterValList[Registers$R0] <- mem_read(RegisterValList[Registers$PC]) + PCOffset
  
  UpdateFlags(RegisterValList[Registers$R0])
}

RTI <- function() {
    abort()
}

RES <- function() {
    abort()
}

AND <- function() {
   RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
   RegisterValList[Registers$R1] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
   ImmFlag <- bitwAnd(bitwShiftR(Instruction, 5), 0x1)
   
    if (ImmFlag) {
      imm5 <- SignExtend(bitwAnd(Instruction, 0x1F), 5)
      RegisterValList[Registers$R0] = bitwAnd(RegisterValList[Registers$R1], imm5)
    } else {
      RegisterValList[Registers$R2] <- bitwAnd(Instruction, 0x7)
      RegisterValList[Registers$R0] <- bitwAnd(RegisterValList[Registers$R1], (RegisterValList[Registers$R2]))
    }
   UpdateFlags(RegisterValList[Registers$R0])
}

NOT <- function() {
   RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
   RegisterValList[Registers$R1] <- bitwAnd(bitwShiftR(Instruction, 6), 0x7)
   
   RegisterValList[Registers$R0] <- bitwNot(RegisterValList[Registers$R1])
   UpdateFlags(RegisterValList[Registers$R0])
   
}

BR <- function() {
  PCOffset <- SignExtend(bitwAnd(Instruction, 0x1FF), 9)
  CondFlag <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
  
  if (bitwAnd(CondFlag, RegisterValList[Registers$COND])) {
    RegisterValList[Registers$COND] <- RegisterValList[Registers$COND] + PCOffset
  }
}

JMP <- function() {
  RegisterValList[Registers$R1] <- bitwAnd(bitwShiftR(Instruction, 6), 0x7)
  RegisterValList[Registers$PC] <-  RegisterValList[Registers$R1]
}

JSR <- function() {
  LongFlag <- bitwAnd(bitwShiftR(Instruction, 11), 1)
  RegisterValList[Registers$R7] <- RegisterValList[Registers$PC]
  
  if (LongFlag) {
    LongPCOffset <- SignExtend(bitwAnd(Instruction, 0x7FF), 11)
    RegisterValList[Registers$PC] <- RegisterValList[Registers$PC] + LongPCOffset
  } else {
    RegisterValList[Registers$R1] <- bitwAnd(bitwShiftR(Instruction, 6), 0x7)
    RegisterValList[Registers$PC] <- RegisterValList[Registers$R1]
  }
}

LD <- function() {
   RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
   PCOffset <- SignExtend(bitwAnd(Instruction, 0x1FF), 9)
   RegisterValList[Registers$R0] <- mem_read(RegisterValList[Registers$PC] + PCOffset) 
   UpdateFlags(RegisterValList[Registers$R0])
}

LDR <- function() {
  RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
  RegisterValList[Registers$R1] <- bitwAnd(bitwShiftR(Instruction, 6), 0x7)
  Offset <- SignExtend(bitwAnd(Instruction, 0x3F), 6)
  RegisterValList[Registers$R0] <- mem_read(RegisterValList[Registers$R1] + Offset)
  
  UpdateFlags(RegisterValList[Registers$R0])
}

LEA <- function() {
  RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
  PCOffset <- SignExtend(bitwAnd(Instruction, 0x1FF), 9)
  
  RegisterValList[Registers$R0] <- RegisterValList[Registers$PC] + PCOffset
  UpdateFlags(RegisterValList[Registers$R0])
}

ST <- function() {
  RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
  PCOffset <- SignExtend(bitwAnd(Instruction, 0x1FF), 9)
  mem_write(RegisterValList[Registers$PC] + PCOffset, RegisterValList[Registers$R0] )
}

STI <- function() {
  RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
  PCOffset <- SignExtend(bitwAnd(Instruction, 0x1FF), 9)
  mem_write(RegisterValList[Registers$PC] + PCOffset, RegisterValList[Registers$R0] )
}


STR <- function() {
    RegisterValList[Registers$R0] <- bitwAnd(bitwShiftR(Instruction, 9), 0x7)
    RegisterValList[Registers$R1] <- bitwAnd(bitwShiftR(Instruction, 6), 0x7)
    Offset <- SignExtend(bitwAnd(Instruction, 0x3F), 6)
    mem_write(RegisterValList[Registers$R1] + PCOffset, RegisterValList[Registers$R0] )
}

```

# Trap Routines

```{r}
TrapCodes <- list(
    TRAP_GETC   = 0x20,
    TRAP_OUT    = 0x21,
    TRAP_PUTS   = 0x22,
    TRAP_IN     = 0x23,  
    TRAP_PUTSP  = 0x24,
    TRAP_HALT   = 0x25
)

TRAP <- function(instr) {
  switch (bitwAnd(instr, 0xFF),
     TRAP_GETC = GETC(),
     TRAP_OUT = OUT(),
     TRAP_PUTS = PUTS(),
     TRAP_IN = IN(),
     TRAP_PUTSP = PUTSP(),
     TRAP_HALT = HALT(),
  )
}

PUTS() <- function() {
  c <- MemoryList[RegisterValList[Registers$R0]]:length(MemoryList)
  idx <- 1
  
  while (c[idx] != 0) {
    cat(rawToChar(as.raw(c[idx])), sep = "")
    idx <- idx + 1
  }
  flush.console()
}

GETC <- function() {
  RegisterValList[Registers$R0] <- readLines(con = stdin(), n = 1)
}


```

